A Simple Minibatch Acceptance Test for MCMC

### 2016-05-18 nips_2016.tex
Updated half of section 3, the first four pages of the writeup look ok, but I need to fix the last part of section 3

In this section, we introduce our new Metropolis-Hastings test and show how to use it to achieve
sufficient mixing and/or to reach better posterior modes. To start, we introduce two key values we
will use frequently, $\Delta$ and $\Delta'$:
\begin{equation}\label{eq:deltas}
\Delta = \log \left( \frac{p(\theta')  \prod_{i=1}^N p(x_i \mid \theta')  q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)  q(\theta' \mid
\theta_t)} \right);\;  \Delta' = \log \left( \frac{p(\theta') \prod_{i=1}^n p(x_i \mid \theta') q(\theta_t \mid
\theta')}{p(\theta_t) \prod_{i=1}^n p(x_i \mid \theta_t) q(\theta' \mid
\theta_t)} \right)
\end{equation}
where $p$ and $q$ have similar definitions as in Equation~\ref{eq:traditional}, so the terms here
consider the likelihood ratios in log %vspace. The only difference between $\Delta$ and $\Delta'$ is
that they use $N$ and $n$ terms in the likelihood ratio, respectively. We assume $n \ll N$ (so
$\Delta'$ is substantially faster to compute) and that the $n$ data points $x_i$ represent a random
minibatch of data.

@@

\begin{proof}
We begin by deriving $p(\theta' \mid \theta_t)$. This is equivalent to the probability of proposing
$\theta'$ and then accepting it, so $p(\theta' \mid \theta_t) = q(\theta' \mid \theta_t)g(\Delta).$
Similarly, $p(\theta_t \mid \theta') = q(\theta_t \mid \theta')g(-\Delta).$ Notice that the
probability of accepting a transition from $\theta'$ to $\theta_t$ is $g(-\Delta)$ because this
inverts the fraction inside the logarithm term of $\Delta$.  By assumption, we can expand $g(\Delta)
= \exp(\Delta)g(-\Delta)$ in $p(\theta' \mid \theta_t)$. Doing this, and combining that with the
definition of $p(\theta_t \mid \theta')$, we get
\begin{equation}\label{eq:combined}
g(-\Delta) = \frac{p(\theta' \mid \theta_t)}{q(\theta' \mid \theta_t)\exp(\Delta)} = \frac{p(\theta_t \mid \theta')}{q(\theta_t \mid \theta')}.
\end{equation}
Rearranging terms and expanding $\exp(\Delta)$, we have
\begin{equation}\label{eq:rearrange}
\frac{p(\theta' \mid \theta_t) f(\theta_t) q(\theta' \mid \theta_t)}{q(\theta' \mid \theta_t) f(\theta') q(\theta_t \mid \theta')} = \frac{p(\theta_t \mid \theta')}{ q(\theta_t \mid \theta')}.
\end{equation}
Cancellations result in $f(\theta') p(\theta_t \mid \theta') = f(\theta_t) p(\theta' \mid
\theta_t)$. Thus, detailed balance is satisfied.
\end{proof}

Some straightforward arithmetic shows that the standard Metropolis-Hastings acceptance function
$g(\Delta) = \min\{1, e^\Delta \} = \min\left\{1, \frac{f(\theta')q(\theta_t \mid
\theta')}{f(\theta_t)q(\theta' \mid \theta_t)}\right\}$ satisfies the condition $g(\Delta) =
\exp(\Delta)g(-\Delta)$ and consequently, results in detailed balance.

For our MH test, the key is that we use a different $g$, the logistic function: $g(\Delta) =
(1+\exp(-\Delta))^{-1}$. Again, straightforward arithmetic shows that it satisfies the
condition in Lemma~\ref{lem:detailed_balance}.  The logistic function is nice because we can easily
sample from it using the following procedure. At any iteration with a current parameter $\theta_t$
and a candidate sample $\theta'$, we can compute $\Delta$. Let $u$ be a uniform random variable $u
\sim {\rm Unif}[0,1]$. We accept $\theta'$ if $g(\Delta) > u$, and reject otherwise. This process is
equivalent to sampling a random variable $X$ with cumulative distribution function $F_X(x) = g(x)$,
and accepting if $\Delta > X$ and rejecting otherwise. We can define $X = g^{-1}(u)$ so that its CDF
is the logistic function, which follows because for arbitrary $X = x$, we have
\[
F_X(x) = {\rm Pr}(X \le x) = {\rm Pr}(g^{-1}(u) \le x) = {\rm Pr}(u \le g(x)) = \int_{0}^{g(x)} 1 dx = g(x),
\]
as the density of $u$ is one. Thus, the criteria to accept the candidate $\theta'$ is equivalent to
whether $\Delta > X$. Moreover, $X$ is symmetric about zero, so the acceptance criteria can also be
expressed as $\Delta + X>0$.

With large datasets, $\Delta$ is intractable to compute. An intuitive alternative is to use
$\Delta'$ instead. The following lemma characterizes the distribution of $\Delta'$.

@@

From Lemma~\ref{lem:gaussian}, since $\Delta'$ is a noisy approximation of $\Delta$, the
relationship is expressed as
\begin{equation}\label{eq:relationship}
\Delta' = \Delta + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, \sigma^2),
\end{equation}
for a Gaussian noise term $\varepsilon$.  The extra noise means we can no longer directly perform
the $\Delta + X$ test, so we need to slightly change our acceptance criteria. We can decompose $X$
as
\begin{equation}\label{eq:deconvolution}
X = X_{\rm norm}+X_{\rm corr},
\end{equation}
where the first term has the same distribution as $\varepsilon$ and the second term is a zero-mean
``correction'' term.  These two add (i.e., convolve) to form $X$. The criteria to accept can be rewritten as
\begin{equation}\label{eq:criteria}
\Delta + X = \Delta + X_{\rm norm} + X_{\rm corr} \approx \Delta' + X_{\rm corr} >0.
\end{equation}
The approximation exists in Equation~\ref{eq:criteria} because, while we know the complete
distribution of $X$, the $X_{\rm norm}$ distribution must be estimated from minibatches of data.
Since $X_{\rm norm}$ is known a priori to be a zero-mean Gaussian, this task reduces to estimating
its standard deviation.

It is important to understand why we use Equations~\ref{eq:deconvolution} and~\ref{eq:criteria}. The
deconvolution process allows us to use $\Delta' + X_{\rm corr}$ as a proxy for $\Delta + X$. It
works because if we know the distribution of $\varepsilon$ (from Equation~\ref{eq:relationship},
which again reduces to knowing its standard deviation), then we can pretend that our $\Delta'$ is in
fact $\Delta$. To ``insert'' $\varepsilon$ into our test, we deconvolve $X$ so that one of its
components, $X_{\rm norm}$, plays the part of $\varepsilon$. This raises two immediate tasks: during
a given iteration, with current sample $\theta_t$ and proposed sample $\theta'$, how should we get
the densities of $X_{\rm norm}$ and $X_{\rm corr}$?

\begin{enumerate}[noitemsep]
    \item To estimate the variance of $X_{\rm norm}$, we generate $K$ values of $\Delta'$, each
    using a different random minibatch of $n$ data points (but each using the same $\theta_t$ and
    $\theta'$ parameters). In our experiments (see Section~\ref{sec:experiments}), $K$ is usually
    small, around five, and $n$ is usually a few hundred for the Central Limit Theorem to apply.
    Using these $K$ values, we estimate the empirical standard deviation ${\rm std}(\Delta')$.

    \item Once we have estimated ${\rm std}(\Delta')$, we can determine the distribution of
    $X_{\rm corr}$ from standard deconvolution techniques. These rely on the well-known fact that
    the Fourier transform of a convolution is the product of the Fourier transforms. For our
    experiments, we pre-compute $X_{\rm corr}$ values and store them in matrices so that during MCMC
    runs, we sample a standard uniform random variable and use that to directly obtain $X_{\rm
    corr}$.
\end{enumerate}

>>>>

For our new MH test, we use two key values, $\Delta$ and $\Delta'$:
\begin{equation}\label{eq:deltas}
\Delta = \log \left( \frac{p(\theta')  \prod_{i=1}^N p(x_i \mid \theta')  q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)  q(\theta' \mid
\theta_t)} \right);\;  \Delta' = \log \left( \frac{p(\theta') \prod_{i=1}^n p(x_i \mid \theta') q(\theta_t \mid
\theta')}{p(\theta_t) \prod_{i=1}^n p(x_i \mid \theta_t) q(\theta' \mid
\theta_t)} \right)
\end{equation}
where $p$ and $q$ have similar definitions as in Equation~\ref{eq:traditional}. Thus, $\Delta$ and
$\Delta'$ are the likelihood ratios in log space. Note that the only difference between $\Delta$ and
$\Delta'$ is that they use $N$ and $n$ data terms $x_i$, respectively. We assume $n \ll N$ (so
$\Delta'$ is substantially faster to compute) and that the $n$ data points $x_i$ represent a random
minibatch of data.

\subsection{The Full Data Test}\label{ssec:deltas_full}

@@

\begin{proof}
We begin by deriving $p(\theta' \mid \theta_t)$. This is equivalent to the probability of proposing
$\theta'$ and then accepting it, so $p(\theta' \mid \theta_t) = q(\theta' \mid \theta_t)g(\Delta).$
Similarly, $p(\theta_t \mid \theta') = q(\theta_t \mid \theta')g(-\Delta).$ Notice that the
probability of accepting a transition from $\theta'$ to $\theta_t$ is $g(-\Delta)$ because this
inverts the fraction inside the logarithm term of $\Delta$.  By assumption, we can expand $g(\Delta)
= \exp(\Delta)g(-\Delta)$ in $p(\theta' \mid \theta_t)$. Doing this, and combining the result with
the definition of $p(\theta_t \mid \theta')$, we get
\begin{equation}\label{eq:combined}
g(-\Delta) = \frac{p(\theta' \mid \theta_t)}{q(\theta' \mid \theta_t)\exp(\Delta)} = \frac{p(\theta_t \mid \theta')}{q(\theta_t \mid \theta')}.
\end{equation}
Rearranging terms and expanding $\exp(\Delta)$, we have
\begin{equation}\label{eq:rearrange}
\frac{p(\theta' \mid \theta_t) f(\theta_t) q(\theta' \mid \theta_t)}{q(\theta' \mid \theta_t) f(\theta') q(\theta_t \mid \theta')} = \frac{p(\theta_t \mid \theta')}{ q(\theta_t \mid \theta')}.
\end{equation}
Cancellations result in $f(\theta') p(\theta_t \mid \theta') = f(\theta_t) p(\theta' \mid
\theta_t)$. Thus, detailed balance is satisfied.
\end{proof}

As a sanity check, the standard Metropolis-Hastings acceptance function $g(\Delta) = \min\{1,
e^\Delta \} = \min\left\{1, \frac{f(\theta')q(\theta_t \mid \theta')}{f(\theta_t)q(\theta' \mid
\theta_t)}\right\}$ satisfies the condition $g(\Delta) = \exp(\Delta)g(-\Delta)$.

For our MH test, the key is that we use a different $g$, the logistic function: $g(\Delta) =
(1+\exp(-\Delta))^{-1}$. Straightforward arithmetic shows that it satisfies the condition in
Lemma~\ref{lem:detailed_balance}.  The logistic function is nice because we can easily sample from
it using the following procedure. At any iteration with a current parameter $\theta_t$ and a
candidate sample $\theta'$, we can compute $\Delta$. Let $u$ be a uniform random variable $u \sim
{\rm Unif}[0,1]$. We accept $\theta'$ if $g(\Delta) > u$, and reject otherwise. This process is
equivalent to sampling a random variable $X$ with cumulative distribution function $F_X(x) = g(x)$,
and accepting if $\Delta > X$ and rejecting otherwise. We can define $X = g^{-1}(u)$ so that its CDF
is the logistic function, which follows because for arbitrary $X = x$, we have
\[
F_X(x) = {\rm Pr}(X \le x) = {\rm Pr}(g^{-1}(u) \le x) = {\rm Pr}(u \le g(x)) = \int_{0}^{g(x)} 1 dx = g(x),
\]
as the density of $u$ is one. Thus, the criteria to accept the candidate $\theta'$ is equivalent to
whether $\Delta > X$. Moreover, $X$ is symmetric about zero, so the acceptance criteria can also be
expressed as $\Delta + X>0$.

\subsection{The Minibatch Version}\label{ssec:deltas_minibatch}

With large datasets, $\Delta$ is intractable to compute. An intuitive fix is to instead use the
minibatch version $\Delta'$. The following lemma characterizes the distribution of $\Delta'$.

@@

From Lemma~\ref{lem:gaussian}, since $\Delta'$ is a noisy approximation of $\Delta$, the
relationship is expressed as
\begin{equation}\label{eq:relationship}
\Delta' = \Delta + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, \sigma^2),
\end{equation}
for a Gaussian noise term $\varepsilon$.  The extra noise means we can no longer directly perform
the $\Delta + X$ test, so we need to slightly change our acceptance criteria. Our insight is to
decompose $X$ as
\begin{equation}\label{eq:deconvolution}
X = X_{\rm norm}+X_{\rm corr},
\end{equation}
where we assume $X_{\rm norm}$ is a zero-mean Gaussian and $X_{\rm corr}$ is a zero-mean
``correction'' term.  These two add (i.e., their distributions convolve) to form $X$. The criteria
to accept is now
\begin{equation}\label{eq:criteria}
\Delta + X = \Delta + X_{\rm norm} + X_{\rm corr} \approx \Delta' + X_{\rm corr} >0.
\end{equation}
The approximation exists because $X_{\rm norm}$ is an estimate of $\varepsilon$.

It is important to understand why we use Equations~\ref{eq:deconvolution} and~\ref{eq:criteria}. The
deconvolution process allows us to use $\Delta' + X_{\rm corr}$ as a proxy for $\Delta + X$. It
works because if we know the distribution\footnote{Computing the distributions of $\varepsilon$ and
$X_{\rm norm}$ reduces to finding their standard deviations.} of $\varepsilon$ (from
Equation~\ref{eq:relationship}), then we can pretend that our $\Delta'$ is in fact $\Delta$. To
``insert'' $\varepsilon$ into our test, we deconvolve $X$ so that one of its components, $X_{\rm
norm}$, plays the part of $\varepsilon$. In light of our revised MH test, the previous discussion
raises the question: in an iteration of MCMC with current $\theta_t$ and proposed $\theta'$, how do
we know the distribution of $X_{\rm corr}$? We need the correct distribution to sample $X_{\rm
corr}$ appropriately. This procedure consists of two major steps.

\begin{enumerate}[noitemsep]
    \item We first need to estimate the standard deviation of $X_{\rm norm}$. We generate $K$ values
    of $\Delta'$, each using a different random minibatch of $n$ data points, but each using the
    same $\theta_t$ and $\theta'$. In our experiments (see Section~\ref{sec:experiments}), $K$ is
    usually small, around 5-10, and $n$ is usually a few hundred for the Central Limit Theorem to
    apply.  Using these $K$ values, we use the empirical standard deviation ${\rm std}(\Delta')$.

    \item Once we have ${\rm std}(\Delta')$, we can determine the distribution of
    $X_{\rm corr}$ from standard deconvolution techniques. These rely on the well-known fact that
    the Fourier transform of a convolution is the product of the Fourier transforms. In practice,
    we pre-compute distributions of $X_{\rm corr}$ for different ${\rm std}(\Delta')$ values to
    facilitate the sampling process.
\end{enumerate}
